DATA 621---Business Analytics and Data Mining

### 2020-09-18 HW1/HW1_draft_kellogg.Rmd
amber's edits to final draft

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  pdf_document:
    toc: TRUE
    toc_depth: 4
  html_document: default
---

@@

```{r loadData, include=FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(tidymodels)
library(mice)
library(psych)
library(Hmisc)
library(MASS)
library(dplyr)
library(caret)
library(data.table)

@@

```{r wins, fig.height = 2.5, fig.width = 5, fig.align = 'center'}
avgWins <- mean(train_set$TARGET_WINS)
sdev <- sd(train_set$TARGET_WINS)
l1 <- mean(train_set$TARGET_WINS) - (2 * sdev)
l2 <- mean(train_set$TARGET_WINS) + (2 * sdev)
# histogram of wins
p <- ggplot(train_set, aes(x=TARGET_WINS)) +
  geom_histogram() + 
  geom_vline(aes(xintercept = avgWins),
            color = "red", linetype = "dashed", size = 1)+
  geom_vline(aes(xintercept = l1),
            color="blue", linetype = "dashed", size = 1) +
    geom_vline(aes(xintercept = l2),
            color = "blue", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Wins", x = "Number of wins",
       y = "Number of seasons")
p
```

@@

```{r pctNull, fig.height = 2, fig.width = 5, fig.align = 'center'}
pct_null <- data.frame(do.call("rbind", map(train_set %>% dplyr::select(-INDEX),
                                            ~ mean(is.na(.)))))
colnames(pct_null) <- c('PCT_NULL')
totalNulls <- pct_null %>%
  mutate(VARIABLE = rownames(.)) %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0) %>%
  dplyr::select(VARIABLE, PCT_NULL)
ggplot(totalNulls, aes(x = reorder(VARIABLE, PCT_NULL), y = PCT_NULL,
                       label = round(PCT_NULL, 2))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+
  geom_bar(stat = "identity") +
  ggtitle("Variables with Missing Information") +
  xlab("Statistic") + ylab("Percent Missing") + 
  coord_flip() + expand_limits(y = 1)
```

@@

```{r negFactors, fig.align = 'center'}
negSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Errors' = TEAM_FIELDING_E, 
         'Hits allowed' = TEAM_PITCHING_H, 
         'Strikeouts by batters' = TEAM_BATTING_SO, 
         'Caught stealing' = TEAM_BASERUN_CS, 
         'Walks allowed' = TEAM_PITCHING_BB, 
         'Home Runs allowed' = TEAM_PITCHING_HR)
ggplot(data = negSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of wins with respect to recorded stats",
       x = "", y = "Number of Wins") + facet_wrap(~ STAT, scales = "free")
```

@@

```{r posFactors, fig.align = 'center'}
posSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Base hits' = TEAM_BATTING_H, 
         'Doubles' = TEAM_BATTING_2B, 
         'Triples' = TEAM_BATTING_3B, 
         'Home Runs' = TEAM_BATTING_HR, 
         'Walks' = TEAM_BATTING_BB, 
         'Batters hit' = TEAM_BATTING_HBP,
         'Stolen bases' = TEAM_BASERUN_SB,
         'Double plays' = TEAM_FIELDING_DP,
         'Strikeouts' = TEAM_PITCHING_SO)
ggplot(data = posSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of Wins with Respect to Recorded Stats", x = "",
       y = "Number of Wins") + facet_wrap(~ STAT, scales = "free")
```

@@

```{r summaryInfo, fig.height = 3, fig.width = 5, fig.align = 'center'}
#summary(train_set)
totalOutliers <- data.frame(
  sapply(train_set %>% dplyr::select(-INDEX, -TARGET_WINS, - TEAM_BATTING_HBP), 
       function(x){length(boxplot.stats(x)$out)/nrow(train_set)}))
totalOutliers$VARIABLE_NM <- rownames(totalOutliers)
colnames(totalOutliers) <- c('PCT_OUTLIERS', 'VARIABLE_NM')
ggplot(totalOutliers,
       aes(x = reorder(VARIABLE_NM, PCT_OUTLIERS), y=PCT_OUTLIERS,
           label = round(PCT_OUTLIERS, 3))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+ geom_bar(stat = "identity") +
  ggtitle("Percentage of Outliers") + xlab("Statistic") +
  ylab("Percent of Data that is an Outlier") + coord_flip() +
  expand_limits(y = 0.15)
```

@@

```{r fewOutliers, fig.height = 2, fig.width = 6, fig.align = 'center'}
set1 <- train_set %>%
  dplyr::select('Home Runs by batters' = TEAM_BATTING_HR, 
         'Strikeouts by batters' = TEAM_BATTING_SO,  
         'Home Runs allowed' = TEAM_PITCHING_HR) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals <- ggplot(set1 , aes(x = stat, y = value)) + 
    geom_boxplot() + labs(title = "", x = "", y = "Value") +
  facet_wrap( ~ stat, scales = "free")
vals
```

@@

```{r manyOutliers, fig.height = 2, fig.width = 6, fig.align = 'center'}
set2 <-  train_set %>%
  dplyr::select('Walks by batters' = TEAM_BATTING_BB, 
         'Stolen bases' = TEAM_BASERUN_SB,  
         'Hits allowed' = TEAM_PITCHING_H,
         'Errors' = TEAM_FIELDING_E) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals2 <- ggplot(set2 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title = "",x = "", y = "Value") +
  facet_wrap(~ stat, scales = 'free', ncol = 4)
vals2
```

@@

```{r otherOutliers, fig.height = 3.5, fig.width = 7, fig.align = 'center'}
set3 <-  train_set %>%
  dplyr::select('Base Hits by batters' = TEAM_BATTING_H, 
         'Doubles by batters' = TEAM_BATTING_2B,  
         'Triples by batters' = TEAM_BATTING_3B,
         'Caught stealing' = TEAM_BASERUN_CS,
         'Walks allowed' = TEAM_PITCHING_BB,
         'Strikeouts by pitchers' = TEAM_PITCHING_SO,
         'Double plays' = TEAM_FIELDING_DP) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals3 <- ggplot(set3, aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title = "", x = "",y = "Value") +
  facet_wrap(~ stat, scales = 'free', ncol = 4)
vals3
```

@@

```{r outliers_1, fig.height = 4, fig.width = 8, fig.align = 'center'}
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to = "Variable", 
               values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```

@@

```{r iqr_removal_SO}
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO,
                                probs = c(0.25), 
                                na.rm = TRUE) - 1.5 * iqr, 
    TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, 
                                probs = c(0.75), 
                                na.rm = TRUE) + 1.5 * iqr
        ) %>%
  dplyr::select(-iqr)
```
```{r iqr_removal_H}
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_H)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_H > quantile(TEAM_PITCHING_H,
                                probs = c(0.25), 
                                na.rm = TRUE) - 1.5 * iqr, 
    TEAM_PITCHING_H < quantile(TEAM_PITCHING_H, 
                                probs = c(0.75), 
                                na.rm = TRUE) + 1.5 * iqr
        ) %>%
  dplyr::select(-iqr)
```

@@

```{r outlier_boxplot, fig.height = 4, fig.width = 8, fig.align = 'center'}
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to ="Variable", 
               values_to="Value") %>%
  ggplot(aes(x=Variable, y=Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```

@@

```{r impute_train}
train_set <- complete(mice(data = train_set, 
                         method = "pmm", 
                         seed = 9450, 
                         print=FALSE), 3)
```

@@

```{r singles}
train_set<- train_set %>%
  dplyr::select(-INDEX) %>%
  mutate_if(is.integer, as.numeric) %>%
  mutate(TEAM_BATTING_1B =
           TEAM_BATTING_H -
           (TEAM_BATTING_2B + TEAM_BATTING_3B + TEAM_BATTING_HR))
```

@@

```{r make_slg_obp}
train_set<- train_set %>%
  mutate(AB = TEAM_BATTING_SO + TEAM_BATTING_BB + TEAM_BATTING_H,
         SLG = (TEAM_BATTING_1B + 2 * TEAM_BATTING_2B + 3 * TEAM_BATTING_3B +
                  4 * TEAM_BATTING_HR) / AB)
```

@@

```{r slugging_compare, fig.height = 2.5, fig.width = 8, fig.align = 'center'}
modified_set <- train_set %>%
  dplyr::select(TARGET_WINS, SLG, TEAM_BATTING_3B, TEAM_BATTING_HR)
ggplot(data = modified_set %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of wins with respect to recorded stats", x = "",
       y = "Number of Wins") +
  facet_wrap(~ STAT, scales = "free")
```

# MODEL TRAINING
## Model 1
### Base Model
We will start with a simple linear model to serve as a baseline. This includes
all variables in the dataset.

```{r baseModel}
lm_reg = lm(data=train_set, TARGET_WINS ~ .)
```

### Enhanced Model
Model 1 includes all variables except Singles, Triples, Base Hits, Strikeouts by
pitchers, Walks allowed, and Batters Hit by Pitch. The final variables were
chosen as a result of backwards `dplyr::selection` based on null hypothesis
testing for non-zero slope.

### Further data transformation
We will use Cook's distance to remove outliers that are influencing the fit of
the model above. We will use a cutoff of \(\frac{4}{N}\).

```{r cooks_distance}
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B + TEAM_BATTING_HR +
          TEAM_BATTING_BB + TEAM_BATTING_SO + TEAM_BASERUN_SB +
            TEAM_BASERUN_CS + TEAM_PITCHING_HR + TEAM_PITCHING_H +
            TEAM_FIELDING_E + TEAM_FIELDING_DP + SLG, data = train_set)
cooks_dis<- cooks.distance(lm_1)
influential<- as.numeric(names(cooks_dis)[(cooks_dis > (4 / nrow(train_set)))])
train_df2<- train_set[-influential, ]
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B + TEAM_BATTING_HR +
            TEAM_BATTING_BB + TEAM_BATTING_SO + TEAM_BASERUN_SB +
            TEAM_BASERUN_CS + TEAM_PITCHING_HR + TEAM_PITCHING_H +
            TEAM_FIELDING_E + TEAM_FIELDING_DP + SLG, data = train_df2)
mod_sum1 = summary(lm_1)  
```
```{r model1_coef}
knitr::kable(summary(lm_1)$coefficients, digits = 3L,
             caption = 'Backwards Model Regression Output')
```

### Coefficient Discussion
First, we are keeping the coefficient for `TEAM_BATTING_HR` since it's p-value
is marginally above the general threshold and knowledge of the game suggests it
is important. There are some counter intuitive results, which is expected given
that baseball is a messy, imprecise game that has evolved over time. To that
end, we should expect some seemingly strange results from algorithmic
regression. We used a combination of domain knowledge and data analysis to
justify retaining features.

@@

```{r model2C+S}
ts2 <- train_set[, -(17:19)]
ts2 <- data.frame(apply(ts2, 2, scale))
ts2$TARGET_WINS <- train_set$TARGET_WINS
```

@@

```{r model3}
stepdata <- subset(train_set, select = -c(AB, SLG, TEAM_BATTING_HBP))
stepdata <- cbind(stepdata, sapply(stepdata[2:length(stepdata)], function(x) x^2))
names(stepdata) <- make.unique(names(stepdata), sep = "_")
model <- lm(TARGET_WINS ~ ., data = stepdata)
cooksd <- cooks.distance(model)
influential <- as.numeric(names(cooksd)[(cooksd > 4*mean(cooksd, na.rm = TRUE))])
stepdata.2 <- stepdata[-c(influential), ]
# Set up repeated k-fold cross-validation
set.seed(525)
train.control <- trainControl(method = "cv", number = 10)
# Train the model
step.model <- train(TARGET_WINS ~ ., data = stepdata.2, method = "lmStepAIC", 
                   trControl = train.control, trace = FALSE)
# Final model 
step.model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_2B + TEAM_BASERUN_SB + 
                  TEAM_BASERUN_CS + TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_FIELDING_E + 
                  TEAM_FIELDING_DP + TEAM_BATTING_BB_1 + TEAM_BATTING_SO_1 + 
                  TEAM_BASERUN_CS_1 + TEAM_PITCHING_H_1 + TEAM_PITCHING_BB_1 + 
                  TEAM_FIELDING_E_1 + TEAM_FIELDING_DP_1 + TEAM_BATTING_1B_1, 
                data = stepdata.2)
# Summary of the model
mod_sum3 <- summary(step.model)
```

@@

# MODEL SELECTION
## Criteria
In order to select the best model to make predictions, we looked at the
following goodness of fit metrics:

@@

## Performance
```{r performance}
# Model accuracy
results <- data.frame(
  Models = c("Model #1: Backwards" , "Model #2: Pairwise", "Model #3: K-fold"),
  R.squared = c(mod_sum1[["r.squared"]], stmr2, mod_sum3[["r.squared"]]), 
  adj.R.squared = c(mod_sum1[["adj.r.squared"]], stmar2,
                    mod_sum3[["adj.r.squared"]]),
  RSME = c(sqrt(mean(lm_1$residuals ^ 2)), stmRMSE,
           sqrt(mean(step.model$residuals ^ 2))),
  AIC = c(AIC(lm_1), AIC(stepModel), AIC(step.model)))
results %>% knitr::kable(digits = 3L, 
                         caption = 'Performance Statistics & Error Measure of Models')
```

It was interesting to compare the three models given that their performance
statistics and error measurements were not significantly different from each other.
**Model #2: Pairwise** was the worst performing model for all of our metrics, and
was eliminated immediately. **Model #3: K-fold** accounts for nearly 45% of the
variation in the dependent variable with the independent variables because its
\(R^2\) is 0.456 and its adjusted \(R^2\) is = 0.450, both of which are
acceptable for a good model. However, **Model #1: Backwards** has the smallest
RSME of all the models. RMSE is a measure of how far off predictions are from
the actual, giving extra weight to outliers through summing the squared
differences. This demonstrates a greater reduction in the randomness by
**Model #1: Backwards**. But its difference from **Model #3: K-fold** is only
0.022. It is hard to judge if that is a significant difference. We thus also
considered AIC. With AIC, it is not the magnitude of the value which is
important, but the difference between the various model AIC's. The magnitude is
a function of the number of observations. Accepted rules of thumb in the
statistical literature are that if the difference is greater than 10, there is
effectively no support for the model with the larger AIC
(Section 2.6, Burnham & Anderson 2002). As a result, **Model #1: Backwards**
is our selected model.

>>>>

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  pdf_document:
    toc: TRUE
    toc_depth: 2
  html_document: default
---

@@

```{r loadData, include=FALSE}
##################### libraries and variables 
library(tidyverse)
library(ggplot2)
library(corrplot)
library(tidymodels)
library(mice)
library(psych)
library(Hmisc)
library(MASS)
library(dplyr)
library(caret)
library(data.table)

@@

```{r wins, fig.height = 2.5, fig.width = 5, fig.align = 'center'}
##################### DATA EXPLORATION

avgWins <- mean(train_set$TARGET_WINS)
sdev <- sd(train_set$TARGET_WINS)
l1 <- mean(train_set$TARGET_WINS) - (2 * sdev)
l2 <- mean(train_set$TARGET_WINS) + (2 * sdev)
# histogram of wins
p <- ggplot(train_set, aes(x=TARGET_WINS)) +
  geom_histogram() + 
  geom_vline(aes(xintercept = avgWins),
            color = "red", linetype = "dashed", size = 1)+
  geom_vline(aes(xintercept = l1),
            color="blue", linetype = "dashed", size = 1) +
    geom_vline(aes(xintercept = l2),
            color = "blue", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Wins", x = "Number of wins",
       y = "Number of seasons")
p
```

@@

```{r pctNull, fig.height = 2, fig.width = 5, fig.align = 'center'}

# percent null for each variable
pct_null <- data.frame(do.call("rbind", map(train_set %>% dplyr::select(-INDEX),
                                            ~ mean(is.na(.)))))
colnames(pct_null) <- c('PCT_NULL')
totalNulls <- pct_null %>%
  mutate(VARIABLE = rownames(.)) %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0) %>%
  dplyr::select(VARIABLE, PCT_NULL)
ggplot(totalNulls, aes(x = reorder(VARIABLE, PCT_NULL), y = PCT_NULL,
                       label = round(PCT_NULL, 2))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+
  geom_bar(stat = "identity") +
  ggtitle("Variables with Missing Information") +
  xlab("Statistic") + ylab("Percent Missing") + 
  coord_flip() + expand_limits(y = 1)
```

@@

```{r negFactors, fig.align = 'center'}
# variables expected to have a negative impact on wins
negSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Errors' = TEAM_FIELDING_E, 
         'Hits allowed' = TEAM_PITCHING_H, 
         'Strikeouts by batters' = TEAM_BATTING_SO, 
         'Caught stealing' = TEAM_BASERUN_CS, 
         'Walks allowed' = TEAM_PITCHING_BB, 
         'Home Runs allowed' = TEAM_PITCHING_HR)
ggplot(data = negSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of wins with respect to recorded stats",
       x = "", y = "Number of Wins") + facet_wrap(~ STAT, scales = "free")
```

@@

```{r posFactors, fig.align = 'center'}
# variables expected to have a positive impact on wins
posSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Base hits' = TEAM_BATTING_H, 
         'Doubles' = TEAM_BATTING_2B, 
         'Triples' = TEAM_BATTING_3B, 
         'Home Runs' = TEAM_BATTING_HR, 
         'Walks' = TEAM_BATTING_BB, 
         'Batters hit' = TEAM_BATTING_HBP,
         'Stolen bases' = TEAM_BASERUN_SB,
         'Double plays' = TEAM_FIELDING_DP,
         'Strikeouts' = TEAM_PITCHING_SO)
ggplot(data = posSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of Wins with Respect to Recorded Stats", x = "",
       y = "Number of Wins") + facet_wrap(~ STAT, scales = "free")
```

@@

```{r summaryInfo, fig.height = 3, fig.width = 5, fig.align = 'center'}
# understanding outliers by variable
totalOutliers <- data.frame(
  sapply(train_set %>% dplyr::select(-INDEX, -TARGET_WINS, - TEAM_BATTING_HBP), 
       function(x){length(boxplot.stats(x)$out)/nrow(train_set)}))
totalOutliers$VARIABLE_NM <- rownames(totalOutliers)
colnames(totalOutliers) <- c('PCT_OUTLIERS', 'VARIABLE_NM')
ggplot(totalOutliers,
       aes(x = reorder(VARIABLE_NM, PCT_OUTLIERS), y=PCT_OUTLIERS,
           label = round(PCT_OUTLIERS, 3))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+ geom_bar(stat = "identity") +
  ggtitle("Percentage of Outliers") + xlab("Statistic") +
  ylab("Percent of Data that is an Outlier") + coord_flip() +
  expand_limits(y = 0.15)
```

@@

```{r fewOutliers, fig.height = 2, fig.width = 6, fig.align = 'center'}
# distributions of variables with few outliers
set1 <- train_set %>%
  dplyr::select('Home Runs by batters' = TEAM_BATTING_HR, 
         'Strikeouts by batters' = TEAM_BATTING_SO,  
         'Home Runs allowed' = TEAM_PITCHING_HR) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals <- ggplot(set1 , aes(x = stat, y = value)) + 
    geom_boxplot() + labs(title = "", x = "", y = "Value") +
  facet_wrap( ~ stat, scales = "free")
vals
```

@@

```{r manyOutliers, fig.height = 2, fig.width = 6, fig.align = 'center'}
# distributions of variables with many outliers
set2 <-  train_set %>%
  dplyr::select('Walks by batters' = TEAM_BATTING_BB, 
         'Stolen bases' = TEAM_BASERUN_SB,  
         'Hits allowed' = TEAM_PITCHING_H,
         'Errors' = TEAM_FIELDING_E) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals2 <- ggplot(set2 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title = "",x = "", y = "Value") +
  facet_wrap(~ stat, scales = 'free', ncol = 4)
vals2
```

@@

```{r otherOutliers, fig.height = 3.5, fig.width = 7, fig.align = 'center'}
# distribution of all other variables 
set3 <-  train_set %>%
  dplyr::select('Base Hits by batters' = TEAM_BATTING_H, 
         'Doubles by batters' = TEAM_BATTING_2B,  
         'Triples by batters' = TEAM_BATTING_3B,
         'Caught stealing' = TEAM_BASERUN_CS,
         'Walks allowed' = TEAM_PITCHING_BB,
         'Strikeouts by pitchers' = TEAM_PITCHING_SO,
         'Double plays' = TEAM_FIELDING_DP) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals3 <- ggplot(set3, aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title = "", x = "",y = "Value") +
  facet_wrap(~ stat, scales = 'free', ncol = 4)
vals3
```

@@

```{r outliers_1, fig.height = 4, fig.width = 8, fig.align = 'center'}

##################### DATA PREPARATION

# visualizing variables pre outlier removal
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to = "Variable", 
               values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```

@@

```{r iqr_removal_SO}
# remove outliers
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO,
                                probs = c(0.25), 
                                na.rm = TRUE) - 1.5 * iqr, 
    TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, 
                                probs = c(0.75), 
                                na.rm = TRUE) + 1.5 * iqr
        ) %>%
  dplyr::select(-iqr)
```
```{r iqr_removal_H}
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_H)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_H > quantile(TEAM_PITCHING_H,
                                probs = c(0.25), 
                                na.rm = TRUE) - 1.5 * iqr, 
    TEAM_PITCHING_H < quantile(TEAM_PITCHING_H, 
                                probs = c(0.75), 
                                na.rm = TRUE) + 1.5 * iqr
        ) %>%
  dplyr::select(-iqr)
```

@@

```{r outlier_boxplot, fig.height = 4, fig.width = 8, fig.align = 'center'}
# visualizing variables post outlier removal
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to ="Variable", 
               values_to="Value") %>%
  ggplot(aes(x=Variable, y=Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```

@@

```{r impute_train}
# imputing values
train_set <- complete(mice(data = train_set, 
                         method = "pmm", 
                         seed = 9450, 
                         print=FALSE), 3)
```

@@

```{r singles}
# defining singles variable 
train_set<- train_set %>%
  dplyr::select(-INDEX) %>%
  mutate_if(is.integer, as.numeric) %>%
  mutate(TEAM_BATTING_1B =
           TEAM_BATTING_H -
           (TEAM_BATTING_2B + TEAM_BATTING_3B + TEAM_BATTING_HR))
```

@@

```{r make_slg_obp}
# defining at bats and slugging variables
train_set<- train_set %>%
  mutate(AB = TEAM_BATTING_SO + TEAM_BATTING_BB + TEAM_BATTING_H,
         SLG = (TEAM_BATTING_1B + 2 * TEAM_BATTING_2B + 3 * TEAM_BATTING_3B +
                  4 * TEAM_BATTING_HR) / AB)
```

@@

```{r slugging_compare, fig.height = 2.5, fig.width = 8, fig.align = 'center'}
# visualizing slugging compared to wins
modified_set <- train_set %>%
  dplyr::select(TARGET_WINS, SLG, TEAM_BATTING_3B, TEAM_BATTING_HR)
ggplot(data = modified_set %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of wins with respect to recorded stats", x = "",
       y = "Number of Wins") +
  facet_wrap(~ STAT, scales = "free")
```

# BUILD MODELS
## Model 1
### Base Model
We will start with a simple linear model to serve as a baseline. This includes
all variables in the dataset.

```{r baseModel}
##################### BUILD MODELS
##### Model 1: Backwards Selection
lm_reg = lm(data=train_set, TARGET_WINS ~ .)

knitr::kable(summary(lm_reg)$coefficients, digits = 3L,
             caption = 'Base Model Regression Output')
```

We can see off the bat that a few variables *exceed* the 0.05 p-value threshold for 
significance: `TEAM_BATTING_H`, `TEAM_PITCHING_H` and `TEAM_PITCHING_BB`. Additionally, 
we see that both `TEAM_BATTING_1B` and `AB` have no coefficients at all! This is likely 
due to the fact that they are colinear with other variables in the model.

### Enhanced Model

We used a combination of domain knowledge and data analysis to justify removing/retaining 
features. First, we are keeping the coefficient for `TEAM_BATTING_HR` since its p-value
is marginally above the general threshold and knowledge of the game suggests it is important. 
We also chose to remove the `TEAM_BATTING_HBP` variable because most of the values are
missing from the dataset.

The remaining variables were chosen as a result of **backwards selection** based on null 
hypothesis testing for non-zero slope. This is done by removing one variable at a
time to determine the minimum number of significant variables needed for prediction.

Altogether, we end up **removing** 7 total variables: `TEAM_BATTING_1B`, `AB`, 
`TEAM_BATTING_3B`, `TEAM_BATTING_H`, `TEAM_PITCHING_BB`, `TEAM_PITCHING_SO`, and
`TEAM_BATTING_1B`.

### Further data transformation
We will use Cook's distance to remove additional outliers that are influencing the fit of
the model above. We will use a cutoff of \(\frac{4}{N}\).

```{r cooks_distance}
# removing outliers via cooks distance
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B + TEAM_BATTING_HR +
          TEAM_BATTING_BB + TEAM_BATTING_SO + TEAM_BASERUN_SB +
            TEAM_BASERUN_CS + TEAM_PITCHING_HR + TEAM_PITCHING_H +
            TEAM_FIELDING_E + TEAM_FIELDING_DP + SLG, data = train_set)
cooks_dis<- cooks.distance(lm_1)
influential<- as.numeric(names(cooks_dis)[(cooks_dis > (4 / nrow(train_set)))])
train_df2<- train_set[-influential, ]
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B + TEAM_BATTING_HR +
            TEAM_BATTING_BB + TEAM_BATTING_SO + TEAM_BASERUN_SB +
            TEAM_BASERUN_CS + TEAM_PITCHING_HR + TEAM_PITCHING_H +
            TEAM_FIELDING_E + TEAM_FIELDING_DP + SLG, data = train_df2)
mod_sum1 = summary(lm_1)  
```

```{r model1_coef}
knitr::kable(summary(lm_1)$coefficients, digits = 3L,
             caption = 'Backwards Model Regression Output')
```

### Coefficient Discussion
There are some counter intuitive results, which is expected given that baseball is a 
messy, imprecise game that has evolved over time. To that end, we should expect some 
seemingly strange results from algorithmic regression.

@@

```{r model2C+S}
##### Model 2: Pairwise Model
# cleaning data for model 2
ts2 <- train_set[, -(17:19)]
ts2 <- data.frame(apply(ts2, 2, scale))
ts2$TARGET_WINS <- train_set$TARGET_WINS
```

@@

```{r model3}

##### Model 3: K-fold Model
stepdata <- subset(train_set, select = -c(AB, SLG, TEAM_BATTING_HBP))
stepdata <- cbind(stepdata, sapply(stepdata[2:length(stepdata)], function(x) x^2))
names(stepdata) <- make.unique(names(stepdata), sep = "_")
model <- lm(TARGET_WINS ~ ., data = stepdata)
cooksd <- cooks.distance(model)
influential <- as.numeric(names(cooksd)[(cooksd > 4*mean(cooksd, na.rm = TRUE))])
stepdata.2 <- stepdata[-c(influential), ]
# Set up repeated k-fold cross-validation
set.seed(525)
train.control <- trainControl(method = "cv", number = 10)
# Train the model
step.model <- train(TARGET_WINS ~ ., data = stepdata.2, method = "lmStepAIC", 
                   trControl = train.control, trace = FALSE)
# Final model 
step.model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_2B + TEAM_BASERUN_SB + 
                  TEAM_BASERUN_CS + TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_FIELDING_E + 
                  TEAM_FIELDING_DP + TEAM_BATTING_BB_1 + TEAM_BATTING_SO_1 + 
                  TEAM_BASERUN_CS_1 + TEAM_PITCHING_H_1 + TEAM_PITCHING_BB_1 + 
                  TEAM_FIELDING_E_1 + TEAM_FIELDING_DP_1 + TEAM_BATTING_1B_1, 
                data = stepdata.2)
# Summary of the model
mod_sum3 <- summary(step.model)
```

@@

# SELECT MODELS
## Criteria
In order to select the best model to make predictions, we looked at the
following goodness of fit metrics:

@@

## Performance
```{r performance}
##################### SELECT MODELS
# Model accuracy
results <- data.frame(
  Models = c("Model #1: Backwards" , "Model #2: Pairwise", "Model #3: K-fold"),
  R.squared = c(mod_sum1[["r.squared"]], stmr2, mod_sum3[["r.squared"]]), 
  adj.R.squared = c(mod_sum1[["adj.r.squared"]], stmar2,
                    mod_sum3[["adj.r.squared"]]),
  RSME = c(sqrt(mean(lm_1$residuals ^ 2)), stmRMSE,
           sqrt(mean(step.model$residuals ^ 2))),
  AIC = c(AIC(lm_1), AIC(stepModel), AIC(step.model)))
results %>% knitr::kable(digits = 3L, 
                         caption = 'Performance Statistics & Error Measure of Models')
```

```{r errors}

# calculate predictions for each model 
mod1_preds <- data.frame(cbind(train_df2$TARGET_WINS, round(mod_sum1$residuals,0), train_df2$TARGET_WINS + round(mod_sum1$residuals,0) ))
colnames(mod1_preds) <- c('TARGET_WINS', 'RESIDUAL', 'PREDICTED')

mod2_preds <- data.frame(cbind(ts2$TARGET_WINS, round(sstm$residuals,0), ts2$TARGET_WINS + round(sstm$residuals,0) ))
colnames(mod2_preds) <- c('TARGET_WINS', 'RESIDUAL', 'PREDICTED')

mod3_preds <- data.frame(cbind(stepdata.2$TARGET_WINS, round(mod_sum3$residuals,0), stepdata.2$TARGET_WINS + round(mod_sum3$residuals,0) ))
colnames(mod3_preds) <- c('TARGET_WINS', 'RESIDUAL', 'PREDICTED')

# summary statistics for residuals
resid_stats <- data.frame(rbind(summary(mod1_preds$RESIDUAL), summary(mod2_preds$RESIDUAL), summary(mod3_preds$RESIDUAL)))
